---
title: "02 Building modules in `SpaDES`"
author: "Alex M. Chubaty"
date: "`r format(Sys.Date(), format='%B %d %Y')`"
output:
  pdf_document:
    number_sections: yes
    toc: yes
  html_document:
    self_contained: no
    toc: yes
vignette: >
  %\VignetteEngine{knitr::knitr}
  %\VignetteIndexEntry{02 Building modules in SpaDES}
  %\VignetteDepends{SpaDES, igraph, ggplot2, RColorBrewer}
  %\VignetteKeyword{module development}
  %\usepackage[utf8]{inputenc}
---

\newpage

# Introduction

## Module overview

Recall that `SpaDES` simulations are event-driven, meaning that different actions are performed on data objects based on the order of scheduled events.
The central design of `SpaDES` promotes modularity, such that collections of related simulation actions can be grouped together as 'modules' and easily reused among multiple simulations.
Strict modularity requires that modules can act independently, without needing to know about other modules.
Thus each `SpaDES` module must explicitly state its input dependencies (data, package, and parameterization requirements), data outputs, as well as provide other useful metadata and documentation for the user.
Upon initialization of a simulation via `simInit`, the dependencies of every module used are examined and evaluated.
If dependency incompatibilities exists, the initialization fails and the user is notified. Also during this initialization, module code is `source`-ed into the global environment, making all module objects and functions available to the user.

Each `SpaDES` module describes the processes or activities that drive simulation state changes via changes to objects stored in the global environment.
Each activity consists of a collection of events which are scheduled depending on the rules of the simulation.
Each event may evaluate or modify a simulation data object (*e.g.*, update the values on a raster map), or perform other operations such as saving and loading data objects, plotting, or scheduling other events.

### Simulation event list

The event queue is stored in a slot in a `simList` simulation object.
Each event is represented by a `data.table` row consisting of the time the event is to occur (`eventTime`), the name of the module from which the event is taken (`moduleName`), and a character string for the programmer-defined event type (`eventType`).
This list is kept sorted by `eventTime`, and events are processed in sequence beginning at the top of the list. 
Completed events are removed from the queue.

### Module events

When a call to the event list is made, the event is processed by the module specified by `moduleName`.
The module code then determines the event type and executes the code for that event.
For each event type within a module: 1) the instructions for what happens for this event get executed; and 2) there is an optional call to `scheduleEvent`, which schedules a future event.
A module can schedule other event types from within the same module, and should not call other modules because this introduces module dependencies, which breaks the drop-in/replace modularity of your simulation model.

### Module event dependencies

Each module schedules its own events (*e.g.*, a 'fire' module may schedule 'burn' events) and only uses its own data objects (or shared global objects).
Modules that behave in this way are independent of one another, which is the preferred way to design and implement modules.
Maintaining strict modularity allows the removal, additon, and replacement of modules without having to rewrite your code.

Module event dependencies complicate the construction of simulation models, and hinder the ability to develop and deploy models with modularity.
If two modules are actually depedent on each others' events, then you should consider whether they really are separate modules or should be merged into a single module.

\newpage

# Module structure

## Module directory structure (`moduleName/`)

A module consists of a collection of files that ...

```
/moduleRepository
  |_ moduleName/
      |_ moduleName.R           # the actual module code file, incl. module metadata
      |_ moduleName.Rmd         # longform documentation and usage info, etc.
      |_ citation.bib           # properly formatted bibtex citation for the module
      |_ LICENSE.txt            # license file describing the allowed usage etc. of the module
      |_ README.txt             # incl. module metadata in addition to version change info, etc.
      |_ moduleName_x.y.z.zip   # zip archive of previous module versions
```

-----

## Module code file (`moduleName.R`)

A `SpaDES` module consists of a single `.R` source file, whose name matches the name of the module.
This file consists of three parts containing the code for:

1. the metadata describing the module and its data dependencies;
2. defining the event types described in the module;
3. functions describing what happens during the processing of each event type.

### Metadata

In order to interact correctly with one another in a simulation, `SpaDES` modules are designed to be aware of their own dependencies and to share this information with the simulation (and to the user).
During simulation initialization, the `.R` file corresponding to each module used is parsed and the modules' metadata stored in the `simList` object.
As part of this initialization step, the dependencies of each module are extracted from the metadata and are checked against the other modules used in the simulation to ensure that all dependencies can be resolved.

#### Defining module metadata: `defineModule()`

Every module requires complete metadata, structured as a named list, and passed as an argument to the `defineModule` function.

**Element name**      | **Description**
----------------------|-------------------------------------------------
**`name`**            | Name of the module as a character string.
**`description`**     | Description of the module as a character string.
**`keywords`**        | Character vector containing a module's keywords.
**`authors`**         | The author(s) of the module as a `person` object.
**`version`**         | The module version as a `numeric_version`. [Semantic versioning](http://semver.org/) is assumed.
**`spatialExtent`**   | Specifies the module's spatial extent as an `Extent` object.
**`timeframe`**       | Specifies the valid timeframe for which the module was designed to simulate.
&nbsp;                | Must be a `POSIXt` object of length `2`, specifying the start and end times.
&nbsp;                | (NOT YET FULLY IMPLEMENTED.)
**`timestep`**        | Describes the time (in seconds) corresponding to `1.0` simulation time units.
**`citation`**        | A citation for the module, as a character string.
**`reqdPkgs`**        | Character vector of R package names to be loaded.
**`parameters`**      | A `data.frame` constructed using `rbind` with `defineParameter`, specifying module parameters,
&nbsp;                | with columns `paramName`, `paramClass`, and `default`.
&nbsp;                | Default values may be overridden by the user by passing a list of parameters to `simInit`.
**`inputObjects`**    | A `data.frame` specifying the object dependencies of the module, with columns
&nbsp;                | `objectName`, `objectClass`, and `other`. For objects that are used within the module
&nbsp;                | as both an input and an output, add the object to each of these `data.frame`s.
**`outputObjects`**   | A `data.frame` specifying the objects output by the module, following the format of
&nbsp;                | `inputObjects`.

```{r module-metadata, eval=FALSE, echo=TRUE}
## sample module metadata for the default `randomLandscapes` module
## NOTE: long lines have been truncated
defineModule(sim, list(
  name="randomLandscapes",
  description="Generate RasterStack of random maps representative of a forest landscape ...",
  keywords=c("random map", "random landscape"),
  authors=c(person(c("Alex", "M"), "Chubaty", email="Alexander.Chubaty@NRCan.gc.ca",
                   role=c("aut", "cre")),
            person(c("Eliot", "J", "B"), "McIntire", email="Eliot.McIntire@NRCan.gc.ca",
                   role=c("aut", "cre"))),
  version=numeric_version("0.2.0"),
  spatialExtent=raster::extent(rep(NA_real_, 4)),
  timeframe=as.POSIXlt(c(NA, NA)),
  timestep=NA_real_,
  citation=list(),
  reqdPkgs=list("raster", "RColorBrewer", "tkrplot", "RandomFields"),
  parameters=rbind(
    defineParameter("stackName", "character", "randomLandscape"),
    defineParameter("nx", "numeric", 100L),
    defineParameter("ny", "numeric", 100L),
    defineParameter("inRAM", "logical", FALSE),
    defineParameter(".plotInitialTime", "numeric", 0),
    defineParameter(".plotInterval", "numeric", 1),
    defineParameter(".saveInitialTime", "numeric", NA_real_),
    defineParameter(".saveInterval", "numeric", NA_real_)),
  inputObjects=data.frame(objectName=character(),
                          objectClass=character(),
                          other=character(), stringsAsFactors=FALSE),
  outputObjects=data.frame(objectName=simGlobals(sim)$stackName,
                           objectClass="RasterStack",
                           other=NA_character_, stringsAsFactors=FALSE)
))
```

#### Defining module parameters: `defineParameter()`

Parameters here differ from input data objects in that the former are intended to be variable across simulation runs, whereas the latter remain constant.
Parameters are often module-specific, where they are only used within the module they are defined, although it may be useful to globally define some parameters that are intended to be used by multiple modules.
Module-specific parameters are specified using `defineParameter` (with `rbind`) within `defineModule` to build a `data.frame` of input parameters.
Global parameters are defined at the simulation level as part of the `simInit` call.

The parameter list in the `simList` object (accessed via `simParams`) may be used to pass named parameter values to modules.
The general structure of this parameter list is `parameters$moduleName$moduleParameter`.
This nested list structure allows passing as many parameters as needed for your simulation.
We suggest passing a list of all the parameters needed for a single module together.

A module's metadata defines default values for module-specific parameters, and these defaults will be used unless overridden by the user.
Default parameter values can overridden by passing values in the parameter list to `simInit`.

```{r passing-params, eval=FALSE, echo=TRUE}
require(SpaDES)

outputPath <- file.path(tempdir(), "simOutputs")
times <- list(start=0.0, stop=20.0)
parameters <- list(.globals=list(stackName="landscape", .outputPath=outputPath,
                                 burnStats="nPixelsBurned"),
                   .progress=list(NA),
                   randomLandscapes=list(nx=100L, ny=100L, inRAM=TRUE),
                   fireSpread=list(nFires=10L, spreadprob=0.225, its=1e6,
                                   persistprob=0, returnInterval=10, startTime=0,
                                  .plotInitialTime=0.1, .plotInterval=10),
                   caribouMovement=list(N=100L, moveInterval=1,
                                        .plotInitialTime=1.01, .plotInterval=1))
modules <- list("randomLandscapes", "fireSpread", "caribouMovement")
path <- system.file("sampleModules", package="SpaDES")

mySim <- simInit(times=times, params=parameters, modules=modules, path=path)
```

### Event types

Each module may contain an arbitrary number of event types.
Each of these event types are defined within the `doEvent.moduleName` call, and are wrapped in a simple `if`/`else` stanza that matches the called event type (NOTE: when several event types are defined, `switch`/case can faster than `if`/`else`).
To keep the `doEvent.moduleName` code block as clear and readable as possible, keep the definitions of each event type minimal, using functions (defined outside of the block) for the details of what is happening for each event.

```{r event-types, echo=TRUE, eval=FALSE}
## sample event type definitions from the default `randomLandscapes` module
doEvent.randomLandscapes <- function(sim, eventTime, eventType, debug=FALSE) {
  if (eventType=="init") {
    # do stuff for this event
    sim <- randomLandscapesInit(sim)

    # schedule the next events
    sim <- scheduleEvent(sim, simParams(sim)$randomLandscapes$.plotInitialTime,
                         "randomLandscapes", "plot")
    sim <- scheduleEvent(sim, simParams(sim)$randomLandscapes$.saveInitialTime,
                         "randomLandscapes", "save")

  } else if (eventType=="plot") {
    # do stuff for this event
    Plot(getGlobal(simGlobals(sim)$stackName))

    # schedule the next event
    sim <- scheduleEvent(sim, simCurrentTime(sim) +
                           simParams(sim)$randomLandscapes$.plotInterval,
                         "randomLandscapes", "plot")
  } else if (eventType=="save") {

    # do stuff for this event
    saveFiles(sim)

    # schedule the next event
    sim <- scheduleEvent(sim, simCurrentTime(sim) +
                           simParams(sim)$randomLandscapes$.saveInterval,
                         "randomLandscapes", "save")

  } else {
    warning(paste("Undefined event type: \'",
                  simEvents(sim)[1, "eventType", with=FALSE],
                  "\' in module \'",
                  simEvents(sim)[1, "moduleName", with=FALSE] , "\'", sep=""))
  }
  return(invisible(sim))
}
```

### Event functions

Event functions should be defined below the `doEvent.moduleName` code block and follow the naming convention `modulenameEventtype()`.
Keep these function definitions as short and clean as possible (you can further modularize your functions by calling additional subroutines).

Functions should get and return global objects, rather than pass them as function arguments.
This mostly allows for function definitions to be simpler, *i.e.*, they just take the one `sim` argument if parameters are passed within the `simInit` call and are needed within the function `getGlobal("object")`.

Note that every module requires an `"init"` event type, which defines the initialization of the module; however, this `init` event need not do a whole lot (*i.e.*, it can be a stub).
As such, the `modulenameInit()` function is required for initiliazation.
Modules may also include `"save"` and `"plot"` events, though these are optional.

```{r event-functions, echo=TRUE, eval=FALSE}
## sample event functions from the default `randomLandscapes` module
randomLandscapesInit <- function(sim) {
  if (is.null(simParams(sim)$randomLandscapes$inRAM)) {
    inMemory <- FALSE
  } else {
    inMemory <- simParams(sim)$randomLandscapes$inRAM
  }
  # Give dimensions of dummy raster
  nx <- simParams(sim)$randomLandscapes$nx
  ny <- simParams(sim)$randomLandscapes$ny
  template <- raster(nrows=ny, ncols=nx, xmn=-nx/2, xmx=nx/2, ymn=-ny/2, ymx=ny/2)
  speedup <- max(1, nx/5e2)
  # Make dummy maps for testing of models
  DEM <- round(gaussMap(template, scale=300, var=0.03, speedup=speedup,
                        inMemory=inMemory), 1)*1000
  forestAge <- round(gaussMap(template, scale=10, var=0.1, speedup=speedup,
                              inMemory=inMemory), 1)*20
  percentPine <- round(gaussMap(template, scale=50, var=1, speedup=speedup,
                                inMemory=inMemory),1)

  # Scale them as needed
  forestAge <- forestAge/maxValue(forestAge)*100
  percentPine <- percentPine/maxValue(percentPine)*100

  # Make layers that are derived from other layers
  habitatQuality <- (DEM+10 + (forestAge+2.5)*10)/100
  habitatQuality <- habitatQuality/maxValue(habitatQuality)

  # Stack them into a single stack and assign to global env
  mapStack <- stack(DEM, forestAge, habitatQuality, percentPine)
  names(mapStack)<-c("DEM", "forestAge", "habitatQuality", "percentPine")

  setColors(mapStack) <- list(DEM=terrain.colors(100),
                              forestAge=brewer.pal(9,"BuGn"),
                              habitatQuality=brewer.pal(8,"Spectral"),
                              percentPine=brewer.pal(9,"Greens"))
  assignGlobal(simGlobals(sim)$stackName, mapStack)
  return(invisible(sim))
}
```

-----

## Module documentation (`moduleName.Rmd`)

Module documentation should be written so that others can use your module.
As such, this documentation should include a high-level overview of the module as well as low-level, detailed descriptions of its usage.

We strongly recommend using RMarkdown (`.Rmd`) for this documentation, as it is easily human-readible while also being machine-readible and easily converted to a wide variety of file formats including `PDF`, `HTML`, and `DOCX`.
Another benefit of using a text-based file format is to facilitate transparent document versioning and history using version control systems like `git`.
For help writing in RMarkdown, see http://rmarkdown.rstudio.com/.

-----

## Citation (`citation.bib`)

Provides a `BibTeX` formatted citation with which the module can be referenced. Most citation management software can read and output `.bib` format.

-----

## License file (`LICENSE.txt`)

This file describes how your module can be redistributed and used by others.
For help selecting a license, see [choosealicense.com](http://choosealicense.com).

-----

## Readme file (`README.txt`)

It is useful to include module metadata in addition to version change information in this file.

-----

## Versioned zip archives (`moduleName_x.y.z.zip`)

Use [semantic versioning](http://semver.org), such that a version number `x.y.z` corresponds to:

- `x` is the **major** version (incremented when you make incompatible API changes)
- `y` is the **minor** version (incremeted when you add functionality in a backwards-compatible manner)
- `z` is the **patch** version (incremented when you make backwards-compatible bug fixes)

\newpage

# Passing data between modules

Modularity in `SpaDES` simulations is maintained by ensuring modules interact indirectly with one another by reading in and modifying objects in the global environment.

## Data types

As you build your modules for your simulation, you can use any of `R`'s data types to store your objects and data.
In particular, matrices (including vectors) and lists work well for this purpose because, as of `R` version `3.1.0`, they are more efficient, reducing your model's memory footprint and speeding up your code's execution.
Other useful datatypes include `Raster*` and `SpatialPoints*` objects (see examples below).

## Global objects

`R` passes copies of objects to functions (instead of using pass-by-reference), so the typical `R` function never acts directly on the global copy of the variable.
Within a function, all operations are generally performed on a local copy of the object, which is then either discarded or returned back to the parent environment upon when the function completes.
This object-copying behaviour has given `R` a reputation of being slow in comparison to languages like `C` and `FORTRAN`.
However, recent improvements to `R` have made it more memory efficient and faster to execute, in part by minimizing the number of object copies made and lazy loading.

One way to reduce object copying and speed up [scoping](http://adv-r.had.co.nz/Functions.html#lexical-scoping) is to work directly on objects in the global environment.
Use the `assign()` and `get()` functions, with `envir=.GlobalEnv` to work with objects directly in the global environment (`SpaDES` provides the shortcuts `assignGlobal()` and `getGlobal()`).
Likewise, to check for the existence of an object in the global environment, use `exists()` (or the `existsGlobal()`).

*NOTE:* Use caution when using the superassignment operator (`<<-`).

## Module object dependencies

As noted above, modules should not depend on one another for event scheduling.
However, it is often useful to develop collections of modules that interact indirectly and are dependent on shared data objects.
Modules can be designed to rely on outputs (data objects) from other modules.
When objects are shared between modules, it is important to state these dependencies explicitly.
To check whether global objects upon which a model depends exist, a call to `checkObject` is made during module initilization.

Note that modules need not be inter-dependent on one another: module B may depend on module A (for example to initialize a data object), without module A depending on module B.

To view the dependencies for a simulation:

```{r depsGraph, eval=TRUE, echo=TRUE, message=FALSE}
require(igraph) # required for plotting dependency graph
require(SpaDES)

times <- list(start=0.0, stop=20)
parameters <- list(.globals=list(stackName="landscape", burnStats="nPixelsBurned"))
modules <- list("randomLandscapes", "fireSpread", "caribouMovement")
path <- system.file("sampleModules", package="SpaDES")

mySim <- simInit(times=times, params=parameters, modules=modules, path=path)

## examine simulation module dependencies
depsEdgeList(mySim, FALSE)          # all object dependency relationships
plot(depsGraph(mySim, TRUE))        # simplified visual representation
```

The output of the `depsEdgeList` is an edgelist describing the relationships between each module.
In the example above, line 2 of the output indicates that the `fireSpread` module relies on an object `landscape` provided by the `caribouMovement` module.
This edgelist is used inside the `depsGraph` function to produce a dependency graph (network), where each module corresponds to a vertex (node), and object dependencies are denoted by the arrows.
When plotting the dependency graph, it's often useful to use the `plot=TRUE` argument to simplify the network by showing no more than one object dependency between modules.

\newpage

# Core modules

There are a number of built-in modules that provide useful functionality such as plotting and saving.
These modules have their names prefaced with a dot (`.`).
Parameter values for core modules can be accessed by using their accessor methods; however, these are simply entries in a nested list exactly like the parameters passed to user modules, and can be accessed directly as such.

These core modules are summarized in the following table and described further below.

**Module Name**   | **Parameterization**                  | **Accessor Methods**
------------------|---------------------------------------|---------------------
**`.checkpoint`** | `file`: character.                    | `simCheckpointFile()`
&nbsp;            | `interval`: numeric (or `NA`).        | `simCheckpointInterval()`
&nbsp;            | &nbsp;                                | &nbsp;
**`.load`**       | `fileList`: character.                | `simFileList()`
&nbsp;            | &nbsp;                                | &nbsp;
**`.plot`**       | `.plotInitialTime`: numeric (or `NA`) | as list item in your module parameters.
&nbsp;            | `.plotInterval`: numeric (or `NA`)    | as list item in your module parameters.
&nbsp;            | &nbsp;                                | &nbsp;
**`.progress`**   | `graphical`: logical (or `NA`).       | `simProgressGraphical()`
&nbsp;            | `interval`: numeric (or `NA`).        | `simProgressInterval()`
&nbsp;            | &nbsp;                                | &nbsp;
**`.save`**       | `.saveInitialTime`: numeric (or `NA`) | as list item in your module parameters.
&nbsp;            | `.saveInterval`: numeric (or `NA`)    | as list item in your module parameters.
&nbsp;            | `.saveObjects`: character.            | as list item in your module parameters.
&nbsp;            | `.savePath`: character.               | as list item in your module parameters.

## `.checkpoint` module

Schedule automatic simulation checkpointing to allow you to resume a previously saved simulation.
All objects in the global environment including the state of the random number generator are saved and can be reloaded.
Checkpoint frequency and filename can be passed as parameters to the simulation object as illustrated below.
By befault, checkpointing is not used unless the `interval` parameter is provided; the simulation checkpoint file (if one exists) is loaded if the `file` parameter is specified.

```{r checkpoints, echo=TRUE, eval=TRUE, message=FALSE}
require(SpaDES)

# initialize a new simulation, setting the checkpoint interval and filename.
times <- list(start=0, stop=30)
parameters <- list(.globals=list(stackName="landscape"),
                   .checkpoint=list(interval=10, file="chkpnt.RData"))
modules <- list("randomLandscapes")
path <- system.file("sampleModules", package="SpaDES")

mySim <- simInit(times=times, params=parameters, modules=modules, path=path)

# retrieve the checkpoint params from the simulation object
simCheckpointFile(mySim)
simCheckpointInterval(mySim)
```

## `.progress` module

Schedule updates to the simulation progress bar.

```{r progress, echo=TRUE, eval=FALSE, message=FALSE}
# initialize a new simulation, setting the progress parameters
mySim <- simInit(times=list(start=0.0, stop=100.0),
                 params=list(.globals=list(stackName="landscape"),
                             .progress=list(graphical=FALSE, interval=10)),
                 modules=list("randomLandscapes"),
                 path=system.file("sampleModules", package="SpaDES"))

# retrieve the checkpoint params from the simulation object
simProgressGraphical(mySim)
simProgressInterval(mySim)
```

## `.load` and `.save` modules

Schedule object save and file load events by passing parameters to the save and load parameters for each module. 
Unlike the core modules above, loading and saving is designed to be scheduled by the user from within a module.

Loading files can be done be passing a simulation parameter called `.load$fileList` which can be a `list` or a `data.frame` and consists minimally of a column called `files` listing the filepaths of the files to be loaded. 
Additional columns can also be provided:

**Column Name** | **Description**
----------------|---------------------------------------------------------------
**`objs`**      | Character string indicating the name of the object once the file is loaded.
**`fun`**       | Character string indicating the function to be used to load the file.
**`intervals`** | Numeric indicating the interval between repeated loading of the same file.
&nbsp;          | This should be `NA` or the column absent if the file is only loaded once.
**`loadTime`**  | Numeric indicating when the file should be loaded. Defaults to `simTime = 0`, but this
&nbsp;          | can be any time. The loading will be scheduled to occur at the `loadTime`.
&nbsp;          | If the same file is to be loaded many times, but not at a regular interval, then there
&nbsp;          | should be separate line, with a unique `loadTime` for each.
**`args`**      | List of lists of named arguments; one list for each loading function.
&nbsp;          | For example, if `raster` is a loading function, `args = list(native = TRUE)`.
&nbsp;          | If there is only one list, then it is assumed to apply to all load attempts and will be
&nbsp;          | repeated for each load function.

Saving objects to file can be done by passing the appropriate arguments as parameters. 
The key values to include are:

**Argument**            | **Description**
------------------------|-------------------------------------------------------
**`.saveInitialTime`**  | Numeric specifying the time at which the first save is scheduled.
**`.saveInterval`**     | Numeric specifying the interval at which objects will be saved.
&nbsp;                  | Used to schedule save events.
**`.saveObjects`**      | Character vector naming the objects to be saved.
**`.savePath`**         | Character giving the path to which the object will be saved.

```{r load-save, echo=TRUE, eval=TRUE, message=FALSE}
# initialize a new simulation, setting the load and save parameters
filelist <- dir(file.path(find.package("SpaDES", quiet=FALSE), "maps"), full.names=TRUE)
mySim <- simInit(times=list(start=0.0, stop=10),
                 params=list(
                   .globals=list(stackName="landscape"),
                   .load=list(fileList=data.frame(files=filelist, stringsAsFactors=FALSE)),
                   randomLandscapes=list(.saveInitialTime=0,
                                         .saveInterval=10,
                                         .saveObjects=c("landscape"),
                                         .savePath=file.path(tempdir(), "output",
                                                             "randomLandscapes"))
                   ),
                 modules=list("caribouMovement"),
                 path=system.file("sampleModules", package="SpaDES")
)

# retrieve the load and save params from the simulation object
simObjectsLoaded(mySim)   # shows what's been loaded
simFileList(mySim)        # returns empty if objects successfully loaded

simParams(mySim)$randomLandscapes$.saveInitialTime
simParams(mySim)$randomLandscapes$.saveInterval
simParams(mySim)$randomLandscapes$.saveObjects
simParams(mySim)$randomLandscapes$.savePath
```

Saving within a module -- scheduling save events.

```{r save-events, echo=TRUE, eval=FALSE, message=FALSE}
### WITHIN A MODULE:

# schedule a recurring save event
nextSave <- simCurrentTime(mySim) + simParams(mySim)$randomLandscapes$.saveInterval
sim <- scheduleEvent(mySim, nextSave, "randomLandscapes", "save")
```

## `.plot` module

Plotting events can be scheduled similarly to save events, by passing module-specific parameters indicating when to schedule the first plot event and how often to rescheduling recurring plot events.

- `.plotInitialTime`: The time at which the first plot is scheduled.
- `.plotInterval`: The interval at which objects will be plotted. Used to schedule plot events.

```{r plotting, echo=TRUE, eval=FALSE, message=FALSE}
# initialize a new simulation, setting the load and save parameters
mySim <- simInit(times=list(start=0.0, stop=100),
                 params=list(.globals=list(stackName="landscape"),
                   randomLandscapes=list(.plotInitialTime=0, .plotInterval=1)
                   ),
                 modules=list("randomLandscapes"),
                 path=system.file("sampleModules", package="SpaDES")
)

# retrieve the plotting params from the simulation object
simParams(mySim)$randomLandscapes$.plotInitialTime
simParams(mySim)$randomLandscapes$.plotInterval
```

```{r plot-events, echo=TRUE, eval=FALSE, message=FALSE}
### WITHIN A MODULE:

# schedule a recurring plot event
nextPlot <- simCurrentTime(mySim) + simParams(mySim)$randomLandscapes$.plotInterval
mySim <- scheduleEvent(mySim, nextPlot, "randomLandscapes", "save")
```

\newpage

# Using existing `SpaDES` modules

One of the primary motivations of developing the `SpaDES` package is to facilitate module reuse.
We provide a few modules with this package as samples for demos and the vignettes, which can be used as a starting point for your own module development.
We also maintain a publically accessible repository of `SpaDES` modules (details below).

Before using other modules, it is up to the modeler (you!) to understand the technical and scientific limitations of these modules.
Although `SpaDES` provides some built-in functionality to ensure different modules will work together, and warn the user if certain module combinations are invalid, when deciding upon which modules to use, be sure to also:

1. Ensure you are using a suitable `R` version;
2. Check package dependencies and ensure you have the necesary packages installed on your system;
3. Check module object dependencies (`simInit` does this);
4. Check module compatibility (*e.g.*, spatial extent, timeframe, etc.) (`simInit` only does a basic version of this).

See the `SpaDES` wiki ([https://github.com/PredictiveEcology/SpaDES/wiki](https://github.com/PredictiveEcology/SpaDES/wiki)) for more detailed information on getting started.

## Included sample modules

Use `system.file("sampleModules", package="SpaDES")` to get the path to the `sampleModules` directory.
To view the contents of a module, use the `openModules` function.
Note that you shouldn't edit the sample modules directly, but save them as a copy in another location.

### Generate random landscapes (`randomLandscapes`)

Generate RasterStack of random maps representative of a forest landscape (DEM, forestAge, habitatQuality, percentPine).
Requires a global simulation parameter `stackName` be set.

```{r randomLandscapes, echo=TRUE, eval=FALSE}
openModules(system.file("sampleModules", package="SpaDES"), "randomLandscapes")
```

### Simulate fire spread on a landscape (`fireSpread`)

Simulate fire ignition and spread on a landscape, where spread probability varies according to percent pine.
Fire size statistics are collected immediately after each burn event.
Requires a global simulation parameter `stackName` be set.

```{r fireSpread, echo=TRUE, eval=FALSE}
openModules(system.file("sampleModules", package="SpaDES"), "fireSpread")
```

### Agent based model of caribou movement (`caribouMovement`)

Simulate caribou movement via correlated random walk.
Requires a RasterStack object whose name is specified by `simGlobals(sim)$stackName`, containing a RasterLayer named `habitatQuality`.

```{r caribouMovement, echo=TRUE, eval=FALSE}
openModules(system.file("sampleModules", package="SpaDES"), "caribouMovement")
```

## `SpaDES` module repository

**Website:** [https://github.com/PredictiveEcology/SpaDES-modules](https://github.com/PredictiveEcology/SpaDES-modules)

Modules from this (or another suitable GitHub repository) can be downloaded using:

```{r download-module, echo=TRUE, eval=FALSE}
downloadModule("moduleName", "path/to/my/modules/directory")
```

\newpage

# Creating new modules

## Module design considerations

As noted above, modules should function as independently as possible so that they retain their modularity.
While it may be useful for modules to exhibit indirect dependence on each other via shared data objects (such as maps), modules should not depend directly on each other via event scheduling.

## Using the module template

Code for new modules can be developed quickly using the template generator function `newModule` which accepts as arguments the name of the new module, a directory path in which to create the new module, and a logical indicating whether to open the newly created module code file for editing.

```{r create-new-module, eval=FALSE, echo=TRUE, message=FALSE}
# create a new module called "randomLandscape" in the "custom-modules" subdirectory
# and open the resulting file immediately for editing.
newModule(name="randomLandscapes", path="custom-modules", open=TRUE)
```

The newly created file can now be modified in the identified sections and customized to your module.
However, it is very important that you do not edit portions of the file outside of the designated sections or your module may fail to work properly within a simulation.

## Contributing modules

After publication you may wish to make your module more easily accessible for others to use (in fact we encourage it).

**Website:** [https://github.com/PredictiveEcology/SpaDES-modules](https://github.com/PredictiveEcology/SpaDES-modules)

### Minimum requirements

All contributed modules should be well documented. 
In addition to providing *complete* metadata within the module's `.R` file, the following components (see details above) should be included:

1. Complete module documentation, such as an Rmarkdown (`.Rmd`) document that can be built as a `PDF`.
2. A `citation.bib` file containing a BibTeX-formatted citation for your module (or for the publication describing your module). This citation will be used by others when citing your module.
3. A `LICENSE` file describing how your module code may be used and modified.
4. A `README` file containing an overview of your module and updates that are included with each version.

Note that `newModule()` produces skeleton templates of these files.
Additionally, `zipModule()` can be used to build the `.zip` file corresponding to a particular version of your module.

### Module submission

To submit your module for inclusion in our GitHub repository, login to GitHub with your account and initiate a pull request to the repository. 
See the [GitHub help documentation](https://help.github.com/articles/creating-a-pull-request/) for more details.
