 ---
title: "03 Plotting with `SpaDES`"
author: "Eliot J. B. McIntire"
date: "`r format(Sys.Date(), format='%B %d %Y')`"
output:
  pdf_document:
    number_sections: yes
    toc: yes
  html_document:
    self_contained: no
    toc: yes
vignette: >
  %\VignetteEngine{knitr::knitr}
  %\VignetteIndexEntry{03 Plotting with SpaDES}
  %\VignetteDepends{SpaDES}
  %\VignetteKeyword{visualization, plotting, mapping}
  %\usepackage[utf8]{inputenc}
---

\newpage

# Plotting in `SpaDES`

Most visualization approaches in R  (*e.g.*, `base`, `RgoogleVis`, `ggplot2`, `rgl`) are not optimized for speed or modularity. Rather, they are designed to build publication-quality or interactive figures that do not need to be redrawn 100s or more times in rapid succession. Simulation modeling is different. Changes to data, maps, objects can happen rapidly and repeatedly. Visualizing this can be critical, especially during the model building phase. Furthermore, no plotting method has been designed for modularity--the ability to do something without being aware of what is already there. 

`SpaDES` comes with a plotting function to fulfill these two gaps. The main plotting function, `Plot` (*i.e.*, with a capital `P`), is built using the grid package. We have specifically built a plotting system that allows for relatively fast plotting of rasters (in the raster package), spatial points, spatial polygons, spatial lines (all within the sp package), ggplot objects, histogram, and igraph (package igraph) objects with the ability to make multi-frame plots without the module (or user) knowing which plots are already plotted. In other words, the main plotting function can handle SpaDES modules, each of which can add plots, without each knowing what the current state of the active plotting device is. This means that plotting can be treated as modular.  Importantly, conventional R plotting (*e.g.*, plot, hist, etc.) can still be used at any time, so you can use the features provided in this package or you can use other R plotting functions without having to relearn a new set of plotting commands. The `Plot` function is therefore intended to be used as a way to interact visually during model development. If fine tuning and customization are desired, other plotting tools may be more suited (*e.g.*, `ggplot2`, or a dedicated GIS program).

To demonstrate plotting, we first load some maps. These maps are randomly generated maps that come with the `SpaDES` package. In the code snippet below, we create the list of files to load, which is every file in the "maps" subdirectory of the package. Then we load that list of files. Because we specified .stackName in the fileList, the loadFiles function will automatically put the individual layers into a `RasterStack`; the individual layers will, therefore, not be available as individual objects within the R environment. If `.stackNames` did not exist, then the individual files would be individual objects.

```{r load_files, eval=TRUE, echo=TRUE, message=FALSE, fig.height=2}
#  Make list of maps from package database to load, and what functions to use to load them
require(magrittr)
require(SpaDES)
fileList <- find.package("SpaDES", lib.loc=getOption("devtools.path"), quiet=FALSE) %>%
       file.path(., "maps") %>% 
         dir(., full.names=TRUE, pattern= "tif") %>%
       data.frame(files=., functions="rasterToMemory", .stackName="landscape",
                  packages="SpaDES", stringsAsFactors=FALSE)

# Load files to memory (using rasterToMemory) and stack them (because .stackName is provided above)
loadFiles(fileList=fileList)
# extract a single one of these rasters
DEM <- landscape$DEM
```

# The `Plot` function
There are several features of `Plot` that are worth highlighting. 

## Layer types
Its main purpose is to plot spatial type objects. Specifically, it currently can plot `RasterLayer`, `RasterStack`, `SpatialPoints*`, `SpatialPolygons*`, and `SpatialLines*` objects. Because `Plot` uses the `grid` package, changing plot parameters is with the `gp=gpar()` designation. See `?gpar` for options.
```{r first_plot, eval=TRUE, echo=TRUE, fig.height=2}
Plot(landscape, new=TRUE)
# make a SpatialPoints object
caribou <- SpatialPoints(coords=cbind(x=runif(1e2,-50,50),y=runif(1e2,-50,50)))
Plot(caribou)
Plot(caribou, addTo="landscape$habitatQuality")

# from SpatialPolygons help file
Sr1 = Polygon(cbind(c(2,4,4,1,2),c(2,3,5,4,2))*20-50)
Sr2 = Polygon(cbind(c(5,4,2,5),c(2,3,2,2))*20-50)

Srs1 = Polygons(list(Sr1), "s1")
Srs2 = Polygons(list(Sr2), "s2")
SpP = SpatialPolygons(list(Srs1,Srs2), 1:2)
Plot(SpP)
Plot(SpP, addTo="landscape$habitatQuality", gp=gpar(lwd=2))

# from SpatialLines help file
l1 = cbind(c(10,2,30),c(30,2,2))
l1a = cbind(l1[,1]+.05,l1[,2]+.05)
l2 = cbind(c(1,20,3),c(10,1.5,1))
Sl1 = Line(l1)
Sl1a = Line(l1a)
Sl2 = Line(l2)
S1 = Lines(list(Sl1, Sl1a), ID="a")
S2 = Lines(list(Sl2), ID="b")
Sl = SpatialLines(list(S1,S2))
Plot(Sl, gp=gpar(col = c("red", "blue"), lwd=2), addTo="landscape$DEM")
```

## gp, gpText, gpAxis, size
These parameters will specify plot parameters that are available with gpar(). `gp` will adjust plot parameters, `gpText` will adjust the title and legend text only, `gpAxis` will adjust the axes only. `size` adjusts point size in a `SpatialPoints` object. Multiple entries can be used, but they must be named list elements and they must match the `...` items to plot. The naming convention used is `RasterStackName.$layerName`, *i.e.*, `landscape$DEM`.

```{r gp_gpAxis_gpText, eval=TRUE, echo=TRUE, fig.height=2}
Plot(caribou, new=TRUE, gpAxis=gpar(cex=0.4), size=1)
Plot(DEM, gpText=gpar(cex=0.4))
```

## Colors
We likely won't want the default colors for every map. There are two ways to change the color of a map: by adding a `colortable` to a `Raster*` object (*e.g.*, using the `setColors` function in `SpaDES`), or using the `cols` argument. Adding a colortable to a `Raster*` will be more persistent (*i.e.*, it will stay with the same color table between calls to `Plot`) than using an argument in the `Plot` function. Every `RasterLayer` can have a `colortable`, which gives the mapping of raster values to colors. If not already set in the file (many `.tif` files and other formats already have their `colortable` set), we can use `setColors(Raster*)` with a named list of hex colours, if a `RasterStack`, or just a vector of hex colors if only a single `RasterLayer`. These can be easily built with the `RColorBrewer` package, with the function `brewer.pal()`, `colorRampPalette`, `heat.colors()`, or just a character vector of colors that will be interpolated.  Note that overplotting will not overplot the legend; in general, overplotting should be used for cases where the maps are compatible with the underlying map layer. See overplotting below.  

`zero.color` is an optional string indicating the color for zero values, when zero is the minimum value, otherwise, it is treated as any other color. Default is NULL. Use "white" or another color if this is not correct. 

`na.color` is an optional string indicating the color for NA values. The default is transparent. Set this value to "white" or another color if this is not correct. See text about `na.color` below after raster values have changed.

```{r set_colors, eval=TRUE, echo=TRUE, fig.height=2}
# can change color palette
Plot(landscape, new=TRUE)
library(RColorBrewer)
setColors(landscape, n=50) <-
              list(DEM=topo.colors(50),
                   forestCover=colorRampPalette(c("blue","orange","purple","red"))(50),
                   forestAge=brewer.pal("Blues", n=8),
                   habitatQuality=brewer.pal(9, "Spectral"),
                   percentPine=brewer.pal("GnBu", n=8))
Plot(landscape[[2:3]])
```

## `visualSqueeze`
Under most circumstances, the plotting regions will be automatically scaled to maximize the area taken by the map layers, minimizing white space, but allowing axes, legends and titles to be visible when they are plotted. In some devices, this automatic scaling is imperfect, so axes or legends may be squished. The `visualSqueeze` argument is an easy way to shrink or grow the plots on the device. The default value is 0.75 representing ~75% of the area. If the plots need to be slightly smaller, this could be set to 0.6; if they can be larger, visualSqueeze could be set to 0.8. 

```{r visualSqueeze, eval=TRUE, echo=TRUE, fig.height=2, dpi=900}
# x axis gets cut off in pdf and html
Plot(DEM, new=TRUE)
Plot(DEM, visualSqueeze=0.6, new=TRUE)
```
 
A key reason why the legends or axes are cut off sometimes is because there is a minimum threshold for font size for readability. So, either visualSqueeze can be set or making a larger device will usually also solve these problems.

## legend and legendRange
`legend` is a logical indicating whether a legend should automatically be added (the default). `legendRange` will allow overriding of the default automatic legend creation. This is useful in situations where, say, an version of a raster at the initial stage of a simulation does not cover the range of the eventual values of that raster. Since the default is to automatically scale the legend to the values contained in the raster at the first time it is drawn, the legend will not cover the whole necessary range. Adding `legendRange=0:20`, where 0:20 covers the expected range of all stages of that raster, will work.
```{r legendRange, eval=TRUE, echo=TRUE, fig.height=2, dpi=900}
Plot(DEM, legendRange=c(0,500), new=TRUE)
```

## zoomExtent
A plot can be made with a custom zoom. This can either be done by cropping the object at in the global environment, or by doing a `Plot` call, passing an extent object via `zoomExtent`. A common way of using this would be to initially `Plot` an object, then use `clickExtent`, assigning the new extent to an object. Then call `Plot(obj, zoomExtent=newZoom)`. This new zoom will persist with rearrangements.
```{r zoomExtent, eval=TRUE, echo=TRUE, fig.height=2, dpi=900}
Plot(DEM, zoomExtent=extent(c(-1, 10, -1, 20)), new=TRUE)
```

## length
For `SpatialLines*` objects, length will give arrow head length, in inches, meaning that the lines will be arrows.


# Modularity
One of the main purposes of the `Plot` function is modularity. The goal is to enable any `SpaDES` module to be able to add a plot to the plotting device, without being aware of what is already in the plotting device. To do this, there is a hidden global variable (a `.spadesArrN` [where N is the device number] object of S4 class, "arrangement") created when a first Plot function is called. This object keeps the layer names, their extents, and whether they were in a RasterStack (and a few other things). So, when a new `Plot` is called, and `new` is used, then it will simply add the new layer. There may not be space on the plot device for this, in which case, everything will be replotted in a new arrangement, but taking the original R objects. This is different than the grid package engine for replotting. That engine was not designed for large numbers of plots to be added to a region; it slows down immensely as the number of plots increases. 

## The `new` argument
There are essentially 3 types of adding that are addressed by this argument, 1) adding a new plot with enough empty space to accommodate the new plot, 2) without this empty space,  and 3) where the device already has a pre-existing plot of the same name. 

### a new name to a device with enough space
The `Plot` function simply adds the new plot in the available empty space.
```{r simple_add, eval=TRUE, echo=TRUE, fig.height=3}
Plot(landscape, new=TRUE)
# can add a new plot to the plotting window
Plot(caribou, new=FALSE, axes=FALSE)
```

### a new name to a device without enough space
The `Plot` function creates a new arrangement, keeping the pre-exising order of plots, and adding the new plots afterwards. The plots will all be a little bit smaller (assuming the device has not changed size), and they will be in different locations on the device. 
```{r add_with_rearrangement, eval=TRUE, echo=TRUE, fig.height=2}
Plot(landscape[[1:4]], new=TRUE)
# can add a new plot to the plotting window
Plot(caribou, new=FALSE, axes=FALSE)
```

### a pre-existing name to a device
The `Plot` function will overplot the new layer in the location as the layer with the same name. If colors in the layer are not transparent, then this will effectively block the previous plot. *This will automatically set legend, title and axes to FALSE*. 
```{r add_with_same_name, eval=TRUE, echo=TRUE, fig.height=2}
Plot(landscape[[1:3]], new=TRUE)
landscape$forestAge[] = ((landscape$forestAge[] + 10) %% 100)
landscape$forestCover[] = ((landscape$forestCover[] +10) %% 30)
# can add a new plot to the plotting window
Plot(landscape[[2:3]], new=FALSE)
# note that zeros are treated as no color by default. If this is not the correct
# behavior, use `zero.color=NULL`
Plot(landscape[[2:3]], new=FALSE, zero.color=NULL)
```

## Mixing Layer Types
Any combination of `RasterStack`, `RasterLayer`, `SpatialPoints*`, and `SpatialPolygons*` objects can be plotted.

```{r mixing_layer_types, eval=TRUE, echo=TRUE, fig.height=2}
Plot(landscape, caribou, DEM, SpP, new=TRUE, axes=TRUE, gp=gpar(cex=0.5))
```

# Plotting Speed
A second main purpose of the `Plot` function is to plot as fast as possible so that visual updates, which may be frequent, take as little time as possible. To do this, several automatic calculations are made upon a call to `Plot`. First, the number of plots is compared to the physical size of the device window. If the layers are `RasterLayer`s, then they are subsampled before plotting, automatically scaled to the number of pixels that would be discernible by the human eye. If the layer is a `SpatialPoints*` object, then a maximum of 10,000 points will be plotted. These defaults can be adjusted by using the `speedup` argument. Broadly, `speedup` is a number >0, where the default is 1. Numbers >1 will plot faster; numbers between 0 and 1 will plot slower. See below for using the `speedup` argument.

## `speedup`
The `speedup` argument is a *relative* speed increase at the cost of resolution if it is >1. If it is between 0 and 1, it will be a relative speed decrease at the gain of resolution. This may be used successfully when the layer texture is particularly coarse, *i.e.*, there are clusters of identical pixels, so subsampling will have little effect. In the examples below, the speedup gains are modest because the Rasters are relatively small (10,000 pixels). This speed gain will be much greater for larger rasters. 

For `SpatialPoints`, the default is to only plot 10,000 points; if there are more than this in the object, then a random sample will be drawn. Speedup is used as the denominator to determine how many to plot, *i.e.*, `10000/speedup`. 

In the example here, the speedup actually slows down plotting because the rasters are already very small. This would not be the case when the original Raster had 1e8 pixels.
```{r speedup, eval=TRUE, echo=TRUE, fig.height=2}
system.time(Plot(landscape, caribou, DEM, new=TRUE))
system.time(Plot(landscape, caribou, DEM, speedup=10, new=TRUE))
# can add a new plot to the plotting window
```

# Overplotting: `addTo`
There are times when it is useful to add a plot to a different plot with a different name. In these cases, the `new` argument will not work. The argument `addTo` will allow plotting of a `RasterLayer` or `SpatialPoints*` object on top of a `RasterLayer`, *that does not share the same name*. This can be useful to see where agents are on a `RasterLayer`, or if there is transparency on a second `RasterLayer`, it could be plotted on top of a first `RasterLayer`.

```{r add, eval=TRUE, echo=TRUE, fig.height=2}
Plot(landscape, new=TRUE)
Plot(caribou, addTo="landscape$forestAge", size=2, axes=F)
```

# Clearing Plot

A device and all the meta-data associated with that device can be cleared 
with the `new=TRUE` argument within the call to `Plot`.
However, sometimes this is not easy enough because the `Plot` call
may be embedded deep within a module. Sometimes, the metadata will prevent
correct plotting because it comes from a previous plot that must be removed.
This function can be used to clear the device and all the associated metadata manually. 
```{r clearPlot, eval=TRUE, echo=TRUE, fig.height=1}
clearPlot()
Plot(caribou)
```

# Using RStudio Plots window
The built-in RStudio plot window is particularly slow. It is recommended to always create a new plotting device whenever real simulations are being done and they will be substantially faster. This may change in a future version of RStudio. Until then, we have created a function, `dev(x)` which will add devices up to the number in the parenthesis, or switch to that device if it is already open. If an RStudio plot has not been called, `dev()` will create a new device outside RStudio. `dev()` can also be used with a number, as in `dev(x=4)` and this function will activate that device, if already exists, or open up all devices required in sequence from current device to `x`. Since RStudio uses 2 and 3 (for the png), a call to `dev(4)` will either create 3 new devices (device numbers 2, 3 and 4 because device number 1 is never used in R), or 1 new device if plotting to RStudio has already occurred. 

# Interacting with plots

## `clickValues`

This can be used to obtain the values on the plotting device at the locations of the mouse clicks. This will work on multiplanel plots. Note that plotting of rasters with the grid package does not allow for partial pixels to be plotted at the edges of the raster. As a result, the edges of pixels may not perfectly line up with the coordinates that they appear with. *Do not rely on exact values when zoomed it*

```{r clickValues, eval=FALSE, echo=TRUE}
Plot(landscape, new=TRUE)
clickValues(3) # click at three locations on the Plot device
```

## `clickExtent`

This can be used like `zoom` for base package plot window. Click two corners of a Plot. 

```{r clickExtent, eval=FALSE, echo=TRUE}
Plot(landscape, new=TRUE)
clickExtent() # click at two locations on the Plot device
```

# ggplot2 objects
`Plot` can also handle gg objects as created with 
`qplot` or `ggplot`.

```{r ggplot, eval=TRUE, echo=TRUE, cache=TRUE}
require(ggplot2)
ggObj <- qplot(rnorm(1e3))
Plot(ggObj)
```

# Replotting or redrawing the entire plot window

`rePlot` can be used to redraw the entire plot window, either in the same frame as 
before (the default) or in a new frame. If in a new frame, the new device will be 
created, then the entire Plot window will be redrawn in the new device. This function 
can also be used to redraw everything after a window gets resized manually and objects become distorted. This cannot be used to "update" a plot with new values of an argument. That will likely be in a future update.

```{r rePlot, eval=FALSE, echo=TRUE, cache=TRUE}
rePlot()
rePlot(4)
```
