---
title: "03 Using `Cache` with `SpaDES`"
author:
  - "Eliot J. B. McIntire"
date: '`r strftime(Sys.Date(), "%B %d %Y")`'
output:
  rmarkdown::html_vignette:
    fig_width: 7
    number_sections: yes
    self_contained: yes
    toc: yes
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{03 Caching SpaDES simulations}
  %\VignetteDepends{archivist, igraph, quickPlot, raster, SpaDES}
  %\VignetteKeyword{cache}
  %\VignetteEncoding{UTF-8}
---

As part of a reproducible work flow, caching of various function calls are a critical component.
Down the road, it is likely that an entire work flow from raw data to publication, decision support, report writing, presentation building etc., could be built and be reproducible anywhere, on demand.
The `reproducible::Cache` function is built to work with any R function. 
However, it becomes very powerful in a `SpaDES` context because we can build large, powerful applications that are transparent and tied to the raw data that may be *many* conceptual steps upstream in the workflow. 
To do this, we have built several customizations within the `SpaDES` package. 
Important to this is dealing correctly with the `simList`, which is an object that has slot that is an environment. But more important are the various tools that can be used at higher levels, *i.e.*, not just for "standard" functions.

# Caching as part of `SpaDES`

Some of the details of the `simList`-specific features of this `Cache` function include:

- The function converts all elements that have an environment as part of their attributes into a format that has no unique environment attribute, using `format` if a function, and `as.list` in the case of the `simList` environment.

- When used within SpaDES modules, `Cache` (capital C) does not require that the argument `cacheRepo` be specified.
If called from inside a SpaDES module, `Cache` will use the `cacheRepo` argument from a call to `cachePath(sim)`, taking the `sim` from the call stack.
Similarly, if no `cacheRepo` argument is specified, then it will use `getOption("spades.cachePath")`, which will, by default, be a temporary location with no persistence between R sessions!
To persist between sessions, use `SpaDES::setPaths()` every session.

In a `SpaDES` context, there are several levels of caching that can be used as part of a reproducible workflow.
Each level can be used to a modeler's advantage; and, all can be -- and are often -- used concurrently.

## At the `spades` level

And entire call to `spades` or `experiment` can be cached.
This will have the effect of eliminating any stochasticity in the model as the output will simply be the cached version of the `simList`.
This is likely most useful in situations where reproducibility is more important than "new" stochasticity (*e.g.*, building decision support systems, apps, final version of a manuscript).

```{r examples, echo=TRUE, message=FALSE}
library(igraph) # for %>%
library(quickPlot)
library(raster)
library(SpaDES)

mySim <- simInit(times = list(start = 0.0, end = 5.0),
                 params = list(.globals = list(stackName = "landscape", burnStats = "testStats"),
                               randomLandscapes = list(.plotInitialTime = NA)),
                 modules = list("randomLandscapes", "fireSpread"),
                 paths = list(modulePath = system.file("sampleModules", package = "SpaDES")))
```

This functionality can be achieved within a `spades` call.

```{r spades}
# compare caching ... run once to create cache
system.time(outSim <- spades(Copy(mySim), cache = TRUE, notOlderThan = Sys.time()))
```

Notice that any visualizations will happen the first time through, but not the subsequent times.

```{r spades-cached}
# vastly faster 2nd time
system.time(outSimCached <- spades(Copy(mySim), cache = TRUE))
all.equal(outSim, outSimCached)
```

## At the `experiment` level

This functionality can be achieved within an experiment call.
This can be done 2 ways, either "internally" through the cache argument.
This will have the effect of caching *each spades call*, or "externally" which will *cache the entire experiment*.
If there are lots of `spades` calls, then the former will be slow as the `simList` will be digested once per `spades` call.

### Using cache argument

```{r experiment-cache}
system.time(sims1 <- experiment(mySim, replicates = 2, cache = TRUE, .plotInitialTime = NA))

# internal -- second time faster
system.time(sims2 <- experiment(mySim, replicates = 2, cache = TRUE, .plotInitialTime = NA))
all.equal(sims1, sims2)
```

### Wrapping `experiment` with `Cache`

Here, the `simList` (and other arguments to experiment) is hashed once, and if it is found to be the same as previous, then the returned list of `simList` objects is recovered. 
This means that even a very large experiment, with many replicates and combinations of parameters and modules can be recovered very quickly. 
Here we show that you can output objects to disk, so the list of `simList` objects doesn't get too big.
Then, when we recover it in the Cached version, all the files are still there, the list of `simList` objects is small, so very fast to recover.

```{r Cache-experiment}
# External
outputs(mySim) <- data.frame(objectName = "landscape")
system.time(sims3 <- Cache(experiment, mySim, replicates = 3, .plotInitialTime = NA,
                           clearSimEnv = TRUE))
```

The second time is way faster. We see the output files in the same location.

```{r Cache-experiment-2}
system.time(sims4 <- Cache(experiment, mySim, replicates = 3, .plotInitialTime = NA,
                           clearSimEnv = TRUE))
all.equal(sims3, sims4)

dir(outputPath(mySim), recursive = TRUE)
```

Notice that speed up can be enormous; in this case ~100 times.

## Module-level caching

If the parameter `.useCache` in the module's metadata is set to `TRUE`, then *every* event in the module will be cached. 
That means that every time that module is called from within a spades or experiment call, `Cache` will be called. 
Only the objects inside the `simList` that correspond to the `inputObjects` or the `outputObjects` from the module metadata will be assessed for caching. 
 
For general use, module-level caching would be mostly useful for modules that have no stochasticity, such as data-preparation modules, GIS modules etc.

In this example, we will use the cache on the randomLandscapes module.
This means that each subsequent call to spades will result in identical outputs from the `randomLandscapes` module (only!).
This would be useful when only one random landscape is needed simply for trying something out, or putting into production code (*e.g.*, publication, decision support, etc.).

```{r module-level, echo=TRUE}
# Module-level
params(mySim)$randomLandscapes$.useCache <- TRUE
system.time(randomSim <- spades(Copy(mySim), .plotInitialTime = NA,
                                notOlderThan = Sys.time(), debug = TRUE))

# vastly faster the second time
system.time(randomSimCached <- spades(Copy(mySim), .plotInitialTime = NA,
                                      debug = TRUE))
```

Test that only layers produced in `randomLandscapes` are identical, not `fireSpread`.

```{r test-module-level}
layers <- list("DEM", "forestAge", "habitatQuality", "percentPine", "Fires")
same <- lapply(layers, function(l) identical(randomSim$landscape[[l]],
                                             randomSimCached$landscape[[l]]))
names(same) <- layers
print(same) # Fires is not same because all non-init events in fireSpread are not cached
```

## Event-level caching

If the parameter `.useCache` in the module's metadata is set to a *character or character vector*, then that or those event(s), identified by their name, will be cached.
That means that every time the event is called from within a `spades` or `experiment` call, `Cache` will be called.
Only the objects inside the `simList` that correspond to the `inputObjects` or the `outputObjects` as defined in the module metadata  will be assessed for caching inputs or outputs, respectively.
The fact that all and only the named `inputObjects` and `outputObjects` are cached and returned may be inefficient (*i.e.*, it may cache more objects than are necessary) for individual events.

Similar to module-level caching, event-level caching would be mostly useful for events that have no stochasticity, such as data-preparation events, GIS events etc.
Here, we don't change the module-level caching for randomLandscapes, but we add to it a cache for only the "init" event for `fireSpread`.

```{r event-level, echo=TRUE}
params(mySim)$fireSpread$.useCache <- "init"
system.time(randomSim <- spades(Copy(mySim), .plotInitialTime = NA,
                                notOlderThan = Sys.time(), debug = TRUE))

# vastly faster the second time
system.time(randomSimCached <- spades(Copy(mySim), .plotInitialTime = NA,
                                      debug = TRUE))
```

## Function-level caching

Any function can be cached using: `Cache(FUN = functionName, ...)`.

This will be a slight change to a function call, such as:
`projectRaster(raster, crs = crs(newRaster))`
to
`Cache(projectRaster, raster, crs = crs(newRaster))`.

```{r function-level, echo=TRUE}
ras <- raster(extent(0, 1e3, 0, 1e3), res = 1)
system.time(map <- Cache(gaussMap, ras, cacheRepo = cachePath(mySim),
                         notOlderThan = Sys.time()))

# vastly faster the second time
system.time(mapCached <- Cache(gaussMap, ras, cacheRepo = cachePath(mySim)))

all.equal(map, mapCached)
```

## Working with the Cache manually

Since the cache is simply an `archivist` repository, all `archivist` functions will work as is.
In addition, there are several helpers in the `reproducible` package, including `showCache`, `keepCache` and `clearCache` that may be useful.
Also, one can access cached items manually (rather than simply rerunning the same `Cache` function again).

```{r manual-cache}
# examine a part of the Cache
showCache(mySim)[tagKey == "function", -c("artifact")]

if (requireNamespace("archivist")) {
  # get the RasterLayer that was produced with the gaussMap function:
  map <- unique(showCache(mySim, userTags = "gaussMap")$artifact) %>%
    archivist::loadFromLocalRepo(repoDir = cachePath(mySim), value = TRUE)
  clearPlot()
  Plot(map)
}
```

# Workflow

In general, we feel that a liberal use of `Cache` will make a re-useable and reproducible work flow.
`shiny` apps can be made, taking advantage of `Cache`. 
Indeed, much of the difficulty in managing data sets and saving them for future use, can be accommodated by caching.
