---
title:  "Plotting `SpaDES` simulation digrams"
author: "Alex M. Chubaty"
date:   "June 26, 2015"
output:
  html_document:
    number_sections: yes
    self_contained: yes
    toc: yes
vignette: >
  %\VignetteEngine{knitr::knitr}
  %\VignetteIndexEntry{03 Plotting with SpaDES}
  %\VignetteDepends{ggplot2, magrittr, RColorBrewer, SpaDES}
  %\VignetteKeyword{visualization, plotting, mapping}
  %\usepackage[utf8]{inputenc}
---

```{r load-packages, echo=FALSE, message=FALSE, warning=FALSE}
devtools::dev_mode(TRUE)

library(DiagrammeR)
library(magrittr)
library(SpaDES)

SpaDES::clearPlot()
```
# Initiate a simulation

```{r simulation, eval=TRUE, message=FALSE, warning=FALSE}
options("spades.nCompleted" = 200)

outputPath <- file.path(tempdir(), "simOutputs")

parameters <- list(.globals=list(stackName="landscape", burnStats="nPixelsBurned"),
                   .progress=list(NA),
                   randomLandscapes=list(
                     nx=100L, ny=100L, inRAM=TRUE,
                     .plotInitialTime=NA, .plotInterval=NA),
                   fireSpread=list(
                     nFires=10L, spreadprob=0.225, its=1e6,
                     persistprob=0, returnInterval=10, startTime=0,
                     .plotInitialTime=NA, .plotInterval=NA),
                   caribouMovement=list(
                     N=100L, moveInterval=1,
                     .plotInitialTime=NA, .plotInterval=NA))

mySim <- simInit(
  times   = list(start=0.0, stop=3, timeunit="year"),
  params  = parameters,
  modules = list("randomLandscapes", "fireSpread", "caribouMovement"),
  objects = list(), 
  path    = system.file("sampleModules", package="SpaDES")
  ) %>% 
  spades
```

# Module diagram

Based on a "network graph" that shows which modules share objects, but does not identify which objects are shared by which modules. This provides a simulation-level dependency diagram.

```{r module-diagram-sim, eval=TRUE}
moduleDiagram(mySim)
```

# Module object diagram

Based on "sequence diagram" that shows which objects are used in which modules.

```{r sequence-diagram-sim, eval=TRUE, fig.height=8, fig.width=10}
objectDiagram(mySim)
```

# Event diagram

Based on "Gantt chart" that shows the sequence of events in each module.

Simulation time is presented on the x-axis, starting at date `startDate`.
Each module appears in a color-coded row, within which each event for that module is displayed corresponding to the sequence of events for that module.
Note that only the start time of the event is meaningful is these figures: the width of the bar associated with a particular module's event corresponds to the module's timestep unit, not the event's "duration".

```{r gantt-chart-sim, eval=TRUE, fig.height=15, fig.width=10}
eventDiagram(mySim, "2005-06-01")
```
